When Forgiveness Backfires: an AI-Mediated Randomised Controlled Experiment on Rejection Sensitivity and Cooperation

### 2024-07-03 article.Rmd
improved intro and discussion

---
title: "Can Exposure To Forgiving AI Foster Cooperative Play?"
author:
  # "Ismail Guennouni, Georgia Koppe, Christoph Korn"

@@

abstract: |

@@

<!-- Trust-based cooperation is an important element of human social interactions, facilitating seamless interpersonal and intergroup relations -->

<!-- <!-- ### Lack of trust and social dysfunction  -->

<!-- Research into the determinants of psychopathology has also linked trust-based constructs to the emergence of mental health disorders [@fonagy_mentalizing_2017]. Many studies have drawn attention to difficulties in social cognition as a crucial concern within personality disorders (See review in @herpertz_social-cognitive_2014). Individuals with these disorders often find it challenging to initiate and sustain social interactions, leading to behaviors that undermine their relationships. Specifically, individuals facing such social cognitive challenges tend to act uncooperatively, which has been identified as a key marker for the severity and presence of  a cluster of Personality Disorder (PD) symptoms [@mulder_relationship_1999]. -->

<!-- <!-- ### Could lack of trust be due to sampling bias for uncaring early caregivers ? -->

<!-- One perspective of how these difficulties in the social domain emerge is through experience with early caregivers. Attachment theory [@bowlby_attachment_1978] posits that the quality of early caregiver relationships shapes an individual's future ability to form secure attachments and trust others. Research strongly suggests a connection between insecure attachment styles (characterized by anxiety or avoidance) and lower levels of trust in relationships. Individuals with higher levels of insecure attachment may recall negative trust-related experiences more easily, report fewer positive trust experiences, and use less constructive coping strategies when trust is broken [@mikulincer_attachment_1998]. Similarly, @fonagy_role_2014 posit that when a young learner grows up surrounded by unreliable communicators, they may adapt by becoming mistrustful of social knowledge. This mistrust serves as a protective mechanism, particularly within environments where caregivers lack benign intention.  -->

<!-- <!-- In their model, drawn on Bowlby’s (1973) attachment theory, Downey and colleagues (1996, 1998) proposed that RS may evolve through early experiences of repeated rejection from primary caregivers, which lead individuals to form the expectation that they are likely to be rejected by others. This expectation is associated with hypervigilance for rejection-relevant cues, which in turn leads to negative emotional responses and maladaptive behavioral reactions. Even if RS is more strongly activated in threatening situations, it is assumed to be a stable disposition, a sort of implicit theory about the self (i.e., rejectable) and others (i.e., rejecters). As such, it can lead to long-term negative effects on individual functioning (e.g., self-regulation, mental health, per- sonality, social cognition) and undermine relationship quality (Gao, Assink, Cipriani, & Lin, 2017).  -->

<!-- <!-- ### Exposure as a way to correct the sampling bias.  -->

<!-- If this adaptive mistrust is the source of social dysfunction, we can ask whether exposing those who exhibit it to cooperative and forgiving interaction partners might correct this bias.  Research in the fields of behavioral economics and psychology has explored how positive social interactions influence trust and cooperation. The use of the repeated trust game (RTG), a well-established experimental approach, has allowed for the analysis of the development of trust through ongoing interactions [@joyce_trust_1995]. In this paradigm, cycles of mutual trust, where each party's trust is reciprocated with trustworthiness, have the effect of enhancing cooperative behaviors and trust levels, even among individuals who are initially inclined to be distrustful [@king-casas_getting_2005]. @fowler_cooperative_2010 studied behavior in social networks interacting in a public goods game and found that cooperative behavior tends to cluster, suggesting that exposure to cooperative peers can lead to more cooperative behavior. Similarly, research on social learning theory [@bandura_social_1977] has long demonstrated that individuals learn and model the behavior of those around them, indicating that if someone is consistently exposed to cooperative and positive individuals, they're likely to emulate this behavior. These insights highlight that engaging with compassionate and forgiving others can be an effective method for mitigating deeply ingrained mistrust. -->

<!-- <!-- ### Describe experiment -->

<!-- In this study, using a randomised control trial design, we assess the effectiveness of a manipulation aimed at repairing the potential breakdown of cooperation in the RTG from a low investment sent by a computerized investor. The manipulation focuses on repeatedly exposing participants to agents that have been designed to have a limited propensity to retaliate to a non-cooperative action by the participants. In order to select a sample that might exhibit difficulties in building relationships based on Trust dues to adverse early experiences, we focused on stratifying participants based on Rejection Sensitivity (RS): a tendency to anxiously expect, readily perceive, and intensely react to rejection. It has been identified as a potential mechanism linking early interpersonal trauma to negative mental health outcomes [@downey_early_1997]. Extensive research demonstrates strong associations between rejection sensitivity and various mental health conditions, including depression, anxiety, personality disorders, and self-harm (for a review, see @gao_associations_2017). However little is known about whether rejection sensitivity is linked to mistrust and difficulties in cooperation in social dilemmas. By startifying participants based on RS, we can directly examine whether the manipulation is particularly effective  for those most vulnerable to trust-related issues. Additionally, comparing outcomes for those with high and low RS will help us determine if the exposure-based manipulation has differential impacts depending on an individual's underlying sensitivity to rejection. The computerized investor in the RTG was programmed to play according to a hidden Markov model estimated from real players' data in prior research. A key aspect of this model is that the actions of the investor depend on a latent "trust state" which reacts dynamically to the trustees' returns simulating real-life trust-building scenarios. Such a model, informed by empirical data and integrating a responsive trust mechanism, represents a novel approach to study interactive behavior in multi-player games whilst keeping a high degree of experimental control. -->

Trust is fundamental to human social interactions, facilitating seamless relations at both interpersonal and intergroup levels.  The study of psychopathology has linked deficits in trust-based constructs to the development of mental health disorders [@fonagy_mentalizing_2017].  Individuals with personality disorders often struggle to form and maintain social connections, a difficulty reflected in uncooperative behaviors – a marker for the severity of PD symptoms [@herpertz_social-cognitive_2014; @mulder_relationship_1999].

One explanation for such social challenges lies in early caregiver experiences. Attachment theory [@bowlby_attachment_1978] suggests that the quality of these relationships shapes our capacity for secure attachments and trust. Individuals with higher levels of insecure attachment may recall negative trust-related experiences more easily, report fewer positive trust experiences, and use less constructive coping strategies when trust is broken [@mikulincer_attachment_1998]. These insecure attachment patterns are often associated with heightened rejection sensitivity (RS), a tendency to anxiously expect, readily perceive, and intensely react to rejection [@downey_early_1997]. RS has been linked to the development of various mental health conditions, including depression, anxiety, personality disorders, and self-harm [@gao_associations_2017]. Given the potential influence of RS on trust-related dynamics, it becomes particularly relevant to our understanding of social dysfunction. Could an ingrained sensitivity to rejection contribute to a mistrust bias? Similarly, learners exposed to unreliable communicators could develop mistrust of social knowledge as a protective strategy [@fonagy_role_2014].

If this adaptive mistrust is the source of social dysfunction, we can ask whether exposing those who exhibit it to cooperative and forgiving interaction partners might correct this bias.  Research in the fields of behavioral economics and psychology has explored how positive social interactions influence trust and cooperation. The use of the repeated trust game (RTG), a well-established experimental approach, has allowed for the analysis of the development of trust through ongoing interactions [@joyce_trust_1995]. In this paradigm, cycles of mutual trust, where each party's trust is reciprocated with trustworthiness, have the effect of enhancing cooperative behaviors and trust levels, even among individuals who are initially inclined to be distrustful [@king-casas_getting_2005]. @fowler_cooperative_2010 studied behavior in social networks interacting in a public goods game and found that cooperative behavior tends to cluster, suggesting that exposure to cooperative peers can lead to more cooperative behavior. Similarly, research on social learning theory [@bandura_social_1977] has long demonstrated that individuals learn and model the behavior of those around them, indicating that if someone is consistently exposed to cooperative and positive individuals, they're likely to emulate this behavior. These insights highlight that engaging with compassionate and forgiving others can be an effective method for mitigating deeply ingrained mistrust.

<!-- In this study, we use a randomized control trial to test whether exposing participants to forgiving and more cooperative co-players can produce more trustworthy behavior and repair potential breakdowns in RTG cooperation. We stratify the sample of participants based on rejection sensitivity (RS). In order to simulate realistic social interaction while maintaining a high degree of experimental control, the computerized investor in the RTG was designed to play using a generative model of how humans play. We fit Hidden markov Models (HMMs) to real players' data from prior studies to build a responsive yet consistent co-player. A key aspect of this agent is that its actions depend on a latent "trust state" which reacts dynamically to the trustees' returns simulating real-life trust-building scenarios. An advantage of having a generative model of behavior is the possibility of controlling different aspects of the agent's strategy such as its general policy, the propensity to cooperate actively or the propensity to retaliate after breakdowns of cooperation. Such generative  model of behavior, informed by empirical data and integrating a responsive trust mechanism, represent a novel approach to study interactive behavior in multi-player games whilst keeping a high degree of experimental control. In this study, we leverage this flexibility to expose participants in between trust games to artificial agents designed with a limited propensity for retaliation, potentially mitigating ingrained mistrust.  In the control condition, the participants play the human like agents throughout.  We would expect those in the manipulation condition to behave more cooperatively and have a lower propensity to retaliate to the co-player's defection. We would also expect those exhibiting higher rejection sensitivity to be more responsive to the pre-programmed change in the agent's cooperativeness and retaliation propensity. -->

In this study, we use a randomized control online experiment to test whether exposing participants with varying levels of RS to forgiving and more cooperative co-players results in more trustworthy behavior and a repair of potential breakdowns in RTG cooperation. In order to simulate realistic social interaction while maintaining a high degree of experimental control, we take a novel paradigmatic approach: We use generative models of how humans play the RTG to design an agent that plays the role of the investor, based on Hidden Markov Models (HMMs) fitted to real players data. A key aspect of these agents is that their actions depend on a latent "trust state" which reacts dynamically to the trustees' returns simulating real-life trust-building scenarios. An advantage of having such a generative model of behavior is the possibility of controlling different aspects of the agent's strategy such as its general policy, the propensity to cooperate actively or the propensity to retaliate after breakdowns of cooperation. In this study, participants were given the role of the trustee. After playing a 15 rounds RTG with a human-like HMM investor, they were randomly assigned to either a Control or Manipulation condition. In the Manipulation condition, participants were exposed over three RTGs to HMM investors designed with a limited propensity for retaliation, potentially mitigating ingrained mistrust. In the Control condition, participants, played three RTGs against the same human-like HMM. After this exposure phase, all participants played another 15 round RTG with a human-like HMM investor.  We hypothesise that those in the manipulation condition would behave more cooperatively and have a lower propensity to retaliate to the co-player's defection after the exposure phase. We also expect those exhibiting higher RS to be more responsive to the pre-programmed change in the agent's cooperativeness and retaliation propensity.

@@

```{r loadCSV}

library(dplyr)
d_finished <- readRDS("../scriptMongo/data/d_finished.csv")

@@

###### get whether high or low RS and score from df_screening

# Load the "df_screening.RData" file
load("data/df_screening.RData")

RS_data <- df_screening %>%  mutate(high_RS = factor(ifelse(RS_score > 15, 1, 0), 
                                    levels = c(0, 1), 
                                    labels = c("low RS", "high RS")))  %>% 
                             dplyr::select(id, MH.f, RS_score, high_RS)

# Find IDs for those who participated in experiment and add RS_score and high_RS as new columns to d_finished.
# Merging d_finished with relevant columns from RS_data based on matching ids
d_finished <- d_finished %>%
  inner_join(RS_data %>% dplyr::select(id, RS_score, high_RS), by = "id")

@@

A total of 206 participants (56% female) were recruited on the Prolific Academic platform (prolific.co). The mean age of participants was 34.6 years, with an 11.9 years standard deviation. Participants were paid a fixed fee of £6 plus a bonus payment dependent on their performance that averaged £0.5. Participants were pre-screened on Prolific using the Rejection Sensitivity Questionnaire (RSQ) to form two similarly sized groups: One with high RS (RSQ score > 15) and the other with low RS (RSQ score < 10).

@@

The experiment had a 2 (Condition: Manipulation or Control) by 2 (RS : high or low) by 2 (Game: Trust-Game Pre Manipulation, Trust-Game Post Manipulation) design, with repeated measures on the third factor (Figure \@ref(fig:HMMPanels).A). Participants within each pre-screened group were randomly assigned to one of the two levels of the first factor. The games were designed and implemented online using Empirica v1 [@almaatouq_empirica_2021].

@@

\begin{landscape}
```{r HMMPanels, fig.cap= "\\small{A: A timeline of the experiment for both conditions. The RTG is played in dyads, with participants always assigned the role of the trustee and the HMM agent that of the investor. The investor is endowed with 20 units at the start of each round. They need to decide how much of that endowment they want to invest with the trustee. The investment is then multiplied by a factor of 3 and sent to the trustee who needs to decide how much of the multiplied investment they want to send back to the investor. The difference between conditions is the type of agents participants are exposed to in the exposure phase. Panels B - D: We construct the artificial investor agent by fitting a three-state hidden Markov model to data of human investors engaged in the 10 round RTG against human trustees. From the fitted HMM, we get the distribution of investments by the human-like agent, conditional on its latent state as shown in Panel B. The fitted HMM also yields the transition probability of the agent to a state on trial t+1 as a function of the net return (difference between the investment sent and the amount received in return) on trial t as shown in Panel C. Each plot in Panel C represents a different starting latent state on trial t, and each line represents the probability of transitioning to a particular state in trial t+1. Panel D shows the transition probabilities of the forgiving HMM agent which are identical to those of the human-like agent, except in the low trust state. Unlike the human-like HMM, the forgiving HMM always transitions out of the low-trust state, and is more likely to end up in a high-trust state.}",include=T, out.width='1\\linewidth', out.height='1\\textheight', fig.align='center'}

@@

The strategy of the computerised investor was modelled on the behavior of human investors in the RTG using an existing dataset consisting of 388  ten round games with the same co-player. Full detail on the datasets used in the Supplementary Information. Using this data, we estimated a hidden Markov model (HMM) on investors' behavior with three latent states. The HMM consists of three outputs. First, a distribution over the possible investments from 0 to 20 for each latent state of the investor (Figure \@ref(fig:HMMPanels).B). These distributions can be interpreted as reflecting "low-trust", "medium-trust", or "high-trust" dispositions. The HMM also specifies how  the investor can move between these latent states.  The probability of these transitions was modelled as a function of their net return (i.e return - investment) in the previous round (see Figure \@ref(fig:HMMPanels).C). Finally, the HMM also provides as an output the initial distribution of latent states for the investor. 
In order to instigate a potential breakdown of trust, thereby allowing us to probe efforts to repair trust, the computerised agent was programmed to provide a low investment on round 12 (pre-manipulation) or round 13 (post-manipulation). On all other rounds, the investor's actions were determined by randomly drawing an investment from the state-conditional distribution, with the state over rounds determined by randomly drawing the next state from the state-transition distribution as determined from the net return on the previous round (disregarding the net return immediately after the pre-programmed low investment rounds). The initial state for the HMM investor in each instance of the game was the "mid-trust" state.

The HMM resulting from fitting the model to human investor play in the RTG (Human-like HMM) isused as a basis to design a forgiving and ultimately more cooperative agent. To achieve that, we changed the transition function of the investor HMM to make it impossible to remain in a low trust state once the agent transitions there. This is done by selecting  parameters of the transition function to make the probability of remaining in the "low-trust" state, when the agent is in that state, effectively nil. The resulting transition function is shown in Figure \ref{fig:HMMPanels}.D. The policies conditional on the latent states and the transition function in the other latent states remain unchanged.

@@

At the start of the experiment, participants provided informed consent and were instructed the study would consist of three phases in which they would face a different other player. Participants were told their goal was to maximise the number of points in all phases. They were not told the number of rounds of each phase. Participants were randomly assigned to either a Control or Manipulation condition. The timeline of the experiment is shown in Figure \@ref(fig:HMMPanels).A.  Phase one ("pre") consisted of a 15 round RTG in which participants took the role of trustee, facing the same investor over all 15 rounds. On each round, after being informed about the amount sent by the investor participants decided how much of the tripled investment to return to the investor, before continuing to the next round. After completing 15 rounds of the RTG, participants rated how cooperative, forgiving they perceived the investor to be, and whether they would like to play with them again (all on a scale from 1 to 10 with 10 being the most positive rating).

Phase 2 ("exposure") consisted of three 7-round RTGs. Participants in the Manipulation condition faced the forgiving HMM investor and rated the agent on the same attributes as in the pre-manipulation phase.  Those in the Control condition faced the same human-like HMM agent as in the "pre" phase and rated each co-player on the same attributes. To keep the experience similar to the "pre" phase, the agent in the control condition was also designed to send a very low investment in round 5 of each of the three games. In the post-manipulation phase ("post"), participants in both conditions were told they would face a new player, and faced the same human-like HMM as in "pre" phase. Participants then completed the Levels of Personality Functioning Scale (LPFS) questionnaire (see the supplement for details). Finally, participants were asked whether they thought the other players were human or computer agents, then debriefed and thanked for their participation.

## Statistical analysis

To explore whether participants behaved differently in the RTG after the manipulation compared to the control group, we model the percentage return (percentage of tripled investment returned to investor) using a linear mixed-effects model with Phase (RTG game pre vs. post-manipulation), Condition (manipulation vs. control), Investment, and RS (High vs Low RS group) as well as their interactions as fixed effects, and player-wise random intercepts and slopes for Phase.

The model was estimated using the `afex` package [@singmann_afex_2022] in R. More complex models with additional random effects could not be estimated reliably, and as such the estimated model can be considered to include the optimal random effects structure [@matuschek_balancing_2017]. A similar process was used to establish the random effects structures of  linear mixed-effects models used to analyse the HMM agent investments as well as the participants' ratings of the co-players. For the $F$-tests, we used the Kenward-Roger approximation to the degrees of freedom, as implemented in the R package "afex". We Z-transform the Investment variable (subtract the overall investment mean and divide by overall standard deviation) as centering is beneficial to interpreting the main effects more easily in the presence of interactions.

```{r ratingsProcess}

# d_finished now has the 'Condition' column with the value equal to expo_opponent for all rounds and rows per playerId

d_finished <- d_finished  %>% mutate( gameNum.f = factor(gameNumber,labels = c("pre", "expo1", "expo2", "expo3","post"),
                                      levels=c("1", "11", "12", "13", "2")),
                                      phase.f = factor(ifelse(gameNumber %in% c("1", "2"), 
                                                       ifelse(gameNumber == "1", "pre", "post"), NA), levels = c("pre", "post")),
                                      investorState.f = factor(investorState, labels= c("unhappy", "neutral", "happy"), levels =c("unhappy", "neutral", "happy"))) %>% 
                              rename(# Before Exposure
                            rating_cooperative_1 = `ratePLayer_trustGame_1.How cooperative do you think the other player is?`,
                            rating_forgiving_1 = `ratePLayer_trustGame_1.How forgiving do you think the other player is?`,
                            rating_playAgain_1 = `ratePLayer_trustGame_1.If given the choice, would you like to play again with the other player?`,

                            # During Exposure
                            rating_cooperative_11 = `ratePLayer_trustGame_11.How cooperative do you think the other player is?`,
                            rating_forgiving_11 = `ratePLayer_trustGame_11.How forgiving do you think the other player is?`,
                            rating_playAgain_11 = `ratePLayer_trustGame_11.If given the choice, would you like to play again with the other player?`,
                            rating_cooperative_12 = `ratePLayer_trustGame_12.How cooperative do you think the other player is?`,
                            rating_forgiving_12 = `ratePLayer_trustGame_12.How forgiving do you think the other player is?`,
                            rating_playAgain_12 = `ratePLayer_trustGame_12.If given the choice, would you like to play again with the other player?`,
                            rating_cooperative_13 = `ratePLayer_trustGame_13.How cooperative do you think the other player is?`,
                            rating_forgiving_13 = `ratePLayer_trustGame_13.How forgiving do you think the other player is?`,
                            rating_playAgain_13 = `ratePLayer_trustGame_13.If given the choice, would you like to play again with the other player?`,

                            # After Exposure
                            rating_cooperative_2 = `ratePLayer_trustGame_2.How cooperative do you think the other player is?`,
                            rating_forgiving_2 = `ratePLayer_trustGame_2.How forgiving do you think the other player is?`,
                            rating_playAgain_2 = `ratePLayer_trustGame_2.If given the choice, would you like to play again with the other player?`,

                            )

@@

# ggplot(mu,aes(x=gameNum.f)) + 
#   geom_bar(aes(y=mean_forgiv, stat = "identity", position = "dodge", color="blue")) + 
#   geom_bar(aes(y=mean_coop, stat = "identity", position = "dodge", color="darkgreen")) +
#   geom_bar(aes(y=mean_playAgain, stat = "identity", position = "dodge", color="red")) +
#   ggtitle("Player ratings after each game") +
#   xlab("Game Number") +
#   theme_minimal() + 
#   theme(legend.position = "bottom")
#   
```

```{r LPFS}
library(dplyr)

# Step 1: Convert LPFS columns to numeric
d_finished <- d_finished %>%
  mutate(across(starts_with("LPFS"), as.numeric))

# Step 2: Calculate means
# Identify LPFS columns
lpfs_cols = grep("^LPFS", names(d_finished), value = TRUE)
lpfs_1_to_6_cols = grep("^LPFS\\.[1-6]$", names(d_finished), value = TRUE)
lpfs_7_to_12_cols = grep("^LPFS\\.(7|8|9|10|11|12)$", names(d_finished), value = TRUE)

# Calculate means using vectorized operations
d_finished <- d_finished %>%
  mutate(
    LPFS_score = rowMeans(dplyr::select(., all_of(lpfs_cols)), na.rm = TRUE),
    LPFS_self_score = rowMeans(dplyr::select(., all_of(lpfs_1_to_6_cols)), na.rm = TRUE),
    LPFS_other_score = rowMeans(dplyr::select(., all_of(lpfs_7_to_12_cols)), na.rm = TRUE)
  )
save(d_finished, file = "data/d_finished.RData")

```{r dfCorr}

library(dplyr)
library(GGally)

df_corr <- d_finished %>% dplyr::select(id, RS_score, LPFS_score, LPFS_other_score,payoffTrust1,payoffTrust2) %>%
  unique()

>>>>

---
title: "When Forgiveness Backfires: an AI-Mediated Randomised Controlled Experiment on Rejection Sensitivity and Cooperation"
author:
  # "Ismail Guennouni, Georgia Koppe, Christoph Korn"

@@

abstract: |
 Can exposure to forgiving behavior improve cooperative outcomes for individuals prone to rejection sensitivity? This randomized controlled experiment explores this question using a novel approach with Hidden Markov Model (HMM)-based AI agents in repeated trust games (RTGs). Participants (N = 206), pre-screened for high or low rejection sensitivity (RS), interacted with either consistently behaving or more forgiving co-players, which were in reality HMM agents. Unexpectedly, exposure to forgiving agents led to reduced cooperation in subsequent interactions. Control group participants maintained consistent behavior. RS levels influenced perceptions of the agent's cooperativeness but did not significantly affect behavioral outcomes, revealing a perception-behavior dissociation. These findings challenge assumptions about fostering cooperation through simple exposure to positive interactions. They also demonstrate the potential of HMM-based AI agents for studying interactive social dynamics, offering a methodological advancement for future research. Our results suggest the need for nuanced interventions that gradually shape expectations and behaviors to foster cooperative outcomes, particularly for individuals with high rejection sensitivity.

@@

Trust is fundamental to human social interactions, facilitating seamless relations at both interpersonal and intergroup levels. The study of psychopathology has linked deficits in trust-based constructs to the development of mental health disorders [@fonagy_mentalizing_2017]. Individuals with personality disorders (PD) often struggle to form and maintain social connections, a difficulty reflected in uncooperative behaviors – a marker for the severity of PD symptoms [@herpertz_social-cognitive_2014; @mulder_relationship_1999].

One explanation for such social challenges lies in early caregiver experiences. Attachment theory [@bowlby_attachment_1978] suggests that the quality of these relationships shapes our capacity for secure attachments and trust. Individuals with higher levels of insecure attachment may recall negative trust-related experiences more easily, report fewer positive trust experiences, and use less constructive coping strategies when trust is broken [@mikulincer_attachment_1998]. These insecure attachment patterns are often associated with heightened rejection sensitivity (RS), a tendency to anxiously expect, readily perceive, and intensely react to rejection [@downey_implications_1996; @downey_early_1997]. RS has been linked to the development of various mental health conditions, including depression, anxiety, personality disorders, and self-harm [@gao_associations_2017]. A recent meta-analysis revealed prosocial behavior and interpersonal trust as two key processes of interpersonal functioning that are markedly impaired in PDs and which are likely to contribute to interpersonal dysfunction in this population [@hepp_prosociality_2022]. The interaction of rejection sensitivity and trust-based constructs has been explored, particularly in Borderline Personality Disorder (BPD). @miano_rejection_2013 and @richetin_emotional_2018 found that rejection sensitivity mediates the relationship between BPD features and lower trust appraisal. However, the interaction between trustworthiness and rejection sensitivity hasn't been studied as extensively in other personality disorders or in non-clinical populations, leaving a gap in our understanding of how these constructs might interplay. Could an ingrained sensitivity to rejection contribute to reduced pro-social behavior such as a failure to show trustworthiness?

If indeed RS contributes to interpersonal dysfunction through failure to show trustworthiness, could exposing those who exhibit it to cooperative and forgiving interaction partners lead to improved interpersonal outcomes? Research in the fields of behavioral economics and psychology has explored how positive social interactions influence cooperation and pro-social behavior. The use of the repeated trust game (RTG), a well-established experimental approach, has allowed for the analysis of the development of trust through ongoing interactions [@joyce_trust_1995]. In this paradigm, cycles of mutual trust, where each party's trust is reciprocated with trustworthiness, have the effect of enhancing cooperative behaviors even among individuals who are initially inclined to be distrustful [@king-casas_getting_2005]. @fowler_cooperative_2010 studied behavior in social networks interacting in a public goods game and found that cooperative behavior tends to cluster, suggesting that exposure to cooperative peers can lead to more cooperative behavior. Similarly, research on social learning theory [@bandura_social_1977] has long demonstrated that individuals learn and model the behavior of those around them, indicating that if someone is consistently exposed to cooperative and positive individuals, they're likely to emulate this behavior. These insights highlight that engaging with compassionate and forgiving others can be an effective method for mitigating deeply ingrained mistrust and potentially fostering trustworthy behavior.

In this study, we use a randomized controlled online experiment to test whether exposing participants with varying RS levels to forgiving and more cooperative co-players results in more trustworthy behavior and a repair of potential breakdowns in RTG cooperation. To simulate realistic social interaction while maintaining a high degree of experimental control, we take a novel paradigmatic approach: We use generative models of how humans play the RTG to design an agent that plays the role of the investor, based on Hidden Markov Models (HMMs) fitted to real players' data. A key aspect of these agents is that their actions depend on a latent "trust state" which reacts dynamically to the trustees' returns, simulating real-life trust-building scenarios. An advantage of having such a generative model of behavior is the possibility of controlling different aspects of the agent's strategy such as its general policy, the propensity to cooperate actively, or the propensity to retaliate after breakdowns of cooperation.

We pre-screened participants for high or low rejection sensitivity (RS) using a validated questionnaire, then assigned them exclusively to the trustee role in a series of trust games. After playing a 15-round RTG with a human-like HMM investor, they were randomly assigned to either a Control or Manipulation condition. In the Manipulation condition, participants were exposed over three RTGs to HMM investors designed with a limited propensity for retaliation, potentially mitigating ingrained mistrust. In the Control condition, participants played three RTGs against the same human-like HMM. After this exposure phase, all participants played another 15-round RTG with a human-like HMM investor. We expect those exhibiting high RS to be more responsive to the pre-programmed change in the agent's cooperativeness and retaliation propensity. We also hypothesize that those in the Manipulation condition would behave more cooperatively and have a lower propensity to retaliate to the co-player's defection after the exposure phase, with those in the higher RS group showing a bigger effect.

@@

```{r loadData}

load("data/d_finished.RData")

@@

# Step 3:  create binary : high or low RS
d_finished <- d_finished %>%
  mutate(high_RS = factor(ifelse(RS_score > 15, 1, 0), 
                                    levels = c(0, 1), 
                                    labels = c("low RS", "high RS")))

@@

A total of 206 participants (56% female) were recruited on the Prolific Academic platform (prolific.co). The required sample size was determined based on an a priori power analysis. The mean age of participants was 34.6 years, with an 11.9 years standard deviation. Participants were paid a fixed fee of £6 plus a bonus payment dependent on their performance that averaged £0.5. A total of 1195 participants were pre-screened on Prolific using the Rejection Sensitivity Questionnaire (RSQ) to finally select two similarly sized groups: One with high RS (RSQ score > 15) and the other with low RS (RSQ score < 10).

@@

The experiment had a 2 (Condition: Manipulation or Control) by 2 (RS : high or low) by 2 (Game: Trust-Game Pre Manipulation, Trust-Game Post Manipulation) design, with repeated measures on the third factor (Figure \@ref(fig:HMMPanels).A). Participants within each pre-screened group were randomly assigned to one of the two levels of the Condition factor. The games were designed and implemented online using Empirica v1 [@almaatouq_empirica_2021]. The planned experiment received approval from the University of Heidelberg's Medical Faculty ethics commission (ID:S-708/2023) and the experiment was performed in accordance with the ethics board guidelines and regulations. All participants provided informed consent prior to their participation.

@@

The computerised investor consisted of a hidden Markov model (HMM) trained on an independent existing behavioral RTG data set of human investors. This data-driven approach thus sought to learn an investor strategy that mimics human-like interactions.  The data set used for training consists of 388 ten round games with the same player (full details can be found in the Supplementary Information). On this data set, the HMM was inferred with three latent states that could be interpreted as reflecting a “low-trust”, a “medium-trust”, and a “high-trust” state. A separate output distribution, that maps each HMM state onto possible investments from 0 to 20 separately, is learned (Figure \@ref(fig:HMMPanels).B). In analogy to the latent states, these distributions can be interpreted as reflecting “low-trust”, “medium-trust”, or “high-trust” dispositions. Finally, the HMM is specified by transition probabilities that describe the transition between states. The probability of these transitions was modelled as a function of their net return (i.e return - investment) in the previous round (see Figure \@ref(fig:HMMPanels).C)). The initial state for the HMM investor in each instance of the game was set to the “mid-trust” state. Details on how the HMM state conditional probabilities and transition functions are specified can be found in the supplement.

In order to instigate a potential breakdown of trust, thereby allowing us to probe efforts to repair it, the computerised agent was programmed to provide a low investment on round 12 (pre-manipulation) and round 13 (post-manipulation). On all other rounds, the investor’s actions were determined by randomly drawing an investment from the state-conditional distribution, with the state over rounds determined by randomly drawing the next state from the state-transition distribution as determined from the net return on the previous round (disregarding the net return immediately after the pre-programmed low investment rounds).

In all phases of the RTG other than the ‘Exposure phase’ (Figure \ref{fig:HMMPanels}.A), participants interacted with this human-like HMM. In the ‘Manipulation’ Condition of the exposure phase, however, the parameters of this HMM were adjusted to design a ‘forgiving’ and ultimately more cooperative agent. To achieve this, we changed the state transition probabilities of the HMM such that it becomes impossible for it to remain in a low trust state, effectively setting the transition probability for remaining in a “low-trust” state to 0. The resulting transition function is shown in Figure \ref{fig:HMMPanels}.D. The policies conditional on the latent states and the transition function in the other latent states remain unchanged.

\begin{landscape}
```{r HMMPanels, fig.cap= "\\small{A: A timeline of the experiment for both conditions. The RTG is played in dyads, with participants always assigned the role of the trustee and the HMM agent that of the investor. The investor is endowed with 20 units at the start of each round. They need to decide how much of that endowment they want to invest with the trustee. The investment is then multiplied by a factor of 3 and sent to the trustee who needs to decide how much of the multiplied investment they want to send back to the investor. The difference between conditions is the type of agents participants are exposed to in the exposure phase. Panels B - D: We construct the artificial investor agent by fitting a three-state hidden Markov model to data of human investors engaged in the 10 round RTG against human trustees. From the fitted HMM, we get the distribution of investments by the human-like agent, conditional on its latent state as shown in Panel B. The fitted HMM also yields the transition probability of the agent to a state on trial t+1 as a function of the net return (difference between the investment sent and the amount received in return) on trial t as shown in Panel C. Each plot in Panel C represents a different starting latent state on trial t, and each line represents the probability of transitioning to a particular state in trial t+1. Panel D shows the transition probabilities of the forgiving HMM agent in the low trust state. Unlike the human-like HMM, the forgiving HMM always transitions out of the low-trust state, and is more likely to end up in a high-trust state, as a proxy for coaxing behavior. Transitions in the medium and high trust states were identical for both agents.}",include=T, out.width='1\\linewidth', out.height='1\\textheight', fig.align='center'}

@@

@@

At the start of the experiment, participants provided informed consent and were instructed the study would consist of three phases in which they would face a different other player. Participants were told their goal was to maximise the number of points in all phases. They were not told the number of rounds of each phase. Participants were randomly assigned to either a Control or Manipulation condition. The timeline of the experiment is shown in Figure \@ref(fig:HMMPanels).A.  Phase one ("pre") consisted of a 15 round RTG in which participants took the role of trustee, facing the same investor over all 15 rounds. On each round, after being informed about the amount sent by the investor participants decided how much of the tripled investment to return to the investor, before continuing to the next round. Phase 2 ("exposure") consisted of three 7-round RTGs. Participants in the Manipulation condition faced the forgiving HMM investor and rated the agent on the same attributes as in the pre-manipulation phase.  Those in the Control condition faced the same human-like HMM agent as in the "pre" phase and rated each co-player on the same attributes. To keep the experience similar to the "pre" phase, the agent in the Control condition was also designed to send a very low investment in round 5 of each of the three games. In the post-manipulation phase ("post"), participants in both conditions faced the same human-like HMM as in "pre" phase.

At the beginning of each game in all three phases, participants were told they would face a new player and had to wait to be paired with an available co-player. This simulated the waiting time in real social interaction tasks. After completing each RTG in each phase, participants rated how cooperative and forgiving they perceived the co-playerr to be, and whether they would like to play with them again (all on a scale from 1 to 10 with 10 being the most positive rating). After completing the three game phases, participants then completed the Levels of Personality Functioning Scale (LPFS) questionnaire (see the supplement for details). Finally, participants were asked whether they thought the other players were human or computer agents, then debriefed and thanked for their participation.

## Statistical Analysis

To explore whether participants behaved differently in the RTG after the manipulation compared to the Control group, we model the percentage return (percentage of tripled investment returned to investor) using a linear mixed-effects model with Phase (RTG game pre vs. post-manipulation), Condition (manipulation vs. Control), Investment, and RS (High vs Low RS group) as well as their interactions as fixed effects, and player-wise random intercepts and slopes for Phase. The full specification of the statistical model can be found in the supplement. To test whether participants behaved differently between conditions in the Exposure phase, we fit a linear mixed effects model to participants returns only in the exposure phase, with Condition (manipulation vs. Control), Investment, and RS (High vs Low RS group) as well as their interactions as fixed effects, and player-wise random intercepts. Finally, to test whether the HMM agent's behavior differed between Phases, Conditions and RS groups, we estimate a linear mixed-effects model of investments sent by the computerised HMM agent with Condition, Phase and RS and their interaction as fixed effects, and a similar random effects structure to the returns model.

The model was estimated using the `afex` package [@singmann_afex_2022] in R. More complex models with additional random effects could not be estimated reliably, and as such the estimated model can be considered to include the optimal random effects structure [@matuschek_balancing_2017]. A similar process was used to establish the random effects structures of linear mixed-effects models used to analyse the HMM agent investments as well as the participants' ratings of the co-players. There is no agreed upon way to calculate effect sizes for mixed effects models. Instead, we will report on testing differences in marginal means. For the $F$-tests, we used the Kenward-Roger approximation to the degrees of freedom, as implemented in the R package "afex". We Z-transform the Investment variable (subtract the overall investment mean and divide by overall standard deviation) as centering is beneficial to interpreting the main effects more easily in the presence of interactions.

@@
